Seasonality in governing dynamics at submesoscales in the Kuroshio Extension

### 2016-06-28 writeup/rocha_etal.tex
Teriz edits.

\begin{abstract}
    Two new high-resolution numerical simulations with embedded tides suggest a
    strong modulation in the governing near-surface dynamics at submesoscales
    (roughly 10-100 km) in the Kuroshio Extension. Consistent with recent studies, deep late Winter mixed layers
    are prone to baroclinic instabilities, and submesoscale turbulence
    prevails in late Winter/early Spring. Inertia-gravity waves are enhanced near
    the surface when the upper ocean is strongly stratified. Thus the summertime
    re-stratification weakens submesoscale turbulence and enhances intertia-gravity
     waves near the surface. In the Kuroshio Extension,
    inertia-gravity waves strongly dominate the surface kinetic energy and
    sea-surface height variance in late Summer/early Fall at submesoscales.
\end{abstract}

@@

\section{The LLC numerical simulations}
We use the output of two latitude-longitude polar cap (LLC)
MITgcm realistic numerical simulations. The outputs
analyzed here, LLC2160 (1/24$^\circ$) and LLC4320 (1/48$^\circ$)  have
90-vertical-levels simulations from a hierarchy of global
forward Massachusetts Institute of Technology general circulation model (MITgcm)
numerical solutions on a LLC grid   \citep{forget_etal2015} that were
spun up from a ECCO2 adjoint-method state estimate. Both simulations were forced by
tides and 6-hourly surface fluxes. The LLC2160
output spans two years from March 2011 to April 2013; the LLC4320 was spun up from
the LLC2160 simulation, spanning one year from September 2011 to September 2012.
The LLC4320 simulation is an extension of the 5-month long output used by
\citet{rocha_etal2016}.

A key aspect of the LLC simulations is that they were forced by
the 16 most significant tidal components.
Because barotropic tides interact with topography and generate internal
tides that project on mesoscales to submesoscales
\citep[e.g., ][]{rocha_etal2016}, tidal forcing fundamentally distinguishes our analysis
from recent modeling studies that investigated upper-ocean dynamics
and its seasonality \citep{sasaki_etal2014,qiu_etal2014}. Details of the LLC simulations
are provided in the supplemental material.

To study seasonal variations in the upper-ocean dynamics, we focus on the northwest
Pacific, in the vicinities of the Kuroshio
Extension, where previous studies have suggested strong submesoscale seasonality
\citep{sasaki_etal2014,qiu_etal2014}.
We analyze a sub-domain of the LLC4320 and LLC2160 simulations of about 2000 km$^2$
spanning 155-175$^\circ$E; 25-40$^\circ$N (figure \ref{fig1}a). The stratification
in this mesoscale-rich subtropical region undergoes a strong seasonal cycle: wintertime
enhancement of small-scale turbulence de-stratifies the upper ocean, yielding mixed layers
as deep as 300m. In late Spring/early Summer enhanced solar radiation and mixed-layer
instabilities re-stratify the upper ocean, yielding mixed layers as shallow as 40m.
Very importantly, the stratification and its seasonality are well-captured by both LLC simulations:
a comparison with Argo climatology shows that both simulations skillfully represent the Kuroshio
Extension stratification and its seasonal variability  (supplemental material).

@@

\section{Statistics of the surface lateral velocity tensor}
To study the seasonality in the surface velocity, we calculate the lateral velocity tensor
\begin{equation}
\nabla_h \bf{u}_h = \left[\begin{matrix} u_x & u_y\\ v_x&v_y \end{matrix}\right] \,.
\end{equation}
The components of the lateral velocity tensor are calculated using a centered
second-order finite differences scheme. We then calculate
the vertical vorticity $\zeta \equiv v_x - u_y$, lateral rate of strain
$\alpha \equiv [(u_x-v_y)^2 + (u_y + v_x)^2]^{1/2}$, and horizontal divergence $\delta \equiv u_x + v_y$.
These diagnostics highlight the submesoscale structures in the flow
\citep[e.g.,][]{capet_etal2008a,shcherbina_etal2013}.

@@

The bulk of vertical vorticity and lateral strain rate are associated with subinertial flows
($T_f = 2\pi/f_0\approx 23.5$h, where $f_0$ is the inertial frequency at
the mean latitude): daily-averaging the velocity fields suppresses superinertial
motions and reduces the RMS
vorticity by less than 40$\%$ and the RMS strain by less than 10$\%$; the seasonal
cycle remains strong (see
dashed lines in figures \ref{fig2}a-b). Most of this seasonal cycle is associated
with submesoscale flows: smoothing the velocity fields
with a Hanning filter with cut-off scale of 100 km dramatically reduces the RMS
vorticity and strain. The reduction in variance is about 80$\%$ in winter,
yielding RMS vorticity and strain roughly consistent with the diagnostics from
AVISO gridded geostrophic velocities (compare blue lines with black lines in figures \ref{fig2}a-b).
The overall picture that emerges is consistent with recent studies: shallow baroclinic
instabilities energize the submesoscales in late Winter, drawing from the available
potential energy stored in deep mixed layers \citep{sasaki_etal2014,callies_etal2015,callies_etal2016}.

The seasonal cycle of the horizontal divergence, however, showcases the complexity
of the upper-ocean annual variability. If submesoscale eddies and fronts dominated
the near-surface variability all year, then the seasonal cycle of horizontal divergence,
vertical vorticity, and lateral strain rate would be in phase \citep[e.g.,][]{sasaki_etal2014}.
While there is a clear wintertime peak in divergence of daily-averaged velocity
(see dashed lines in figure \ref{fig2}c; RMS divergence $\sim0.22 f$ in the 1/48$^\circ$
simulation), the hourly fields show
a stronger enhancement of
lateral divergence in late Summer/early Fall (RMS divergence $\sim0.22 f$ the 1/48$^\circ$
simulation). Because the 1/48$^\circ$ simulation better resolves
smaller-scale submesoscale flows, a secondary RMS divergence peak in winter  is
nearly as strong as the summertime's. Submesoscale fronts and eddies evolve
relatively fast and there is no clear
temporal and spatial scale separation between those motions and inertia-gravity waves
\citep{mcwilliams2016}:
daily-averaging the velocity fields efficiently suppresses the summertime horizontally
divergent flows, but also reduces the wintertime lateral divergence by about 50$\%$.
The vorticity and strain rate (Figure \ref{fig2}a-b) show that most of the lateral divergence is associated
with submesoscale flows: smoothing the velocity fields with a 100km-cut-off
suppresses more than 80$\%$ of the RMS divergence.

\section{Seasonality in governing submesoscale dynamics}
The overall results of figure \ref{fig2}c shows that summertime submesoscale surface
 variability stems from different dynamics than wintertime variability.
 To better understand these differences, we calculate
joint probability distributions (jPDF) of
vorticity-divergence, vorticity-strain, and vorticity-Laplacian of sea-surface height.

The April vorticity-strain jPDF has a shape characteristic of submesoscale turbulence
 (see figures \ref{fig3}f-g). The alignment of vorticity and strain $\alpha \sim \zeta$
 with strong positive skewness are fingerprints of submesoscale fronts
 \citep{shcherbina_etal2013,mcwilliams2016}. The shape of the vorticity-strain
 jPDF is similar for hourly, daily-averaged, and residual velocities, although
 the vorticity skewness reduces 1.4 to 1.13 from hourly to daily-averaged.
 The April vorticity-divergence jPDFs, however, egg-shaped which again highlights
 the strong skewness in vorticity. Thus wintertime submesoscale surface velocity
 is strongly dominated by submesoscale turbulence --- this is true even for
 superinertial submesoscale currents. The hourly velocity fields are significantly
ageostrophic as depicted by the jPDF of vorticity-Laplacian of SSH
(figure \ref{fig3}o); it is only the daily-averaged velocity that is approximately
in geostrophic balance (figure \ref{fig3}l). While there is certainly much to be
learned using quasi-geostrophic models to study submesoscale turbulence, claims that
the submesoscale turbulence is dominantly in geostrophic balance
are overstated \citep{sasaki_etal2014,callies_etal2016}; only the daily-averaged fields
are largely
in  geostrophic balance (the jPDF of daily-averaged vorticity vs. Laplacian of SSH
is an ellipse with large eccentricity with main axis tilted by 45$^\circ$; figure \ref{fig3}n).

The October vorticity-strain jPDF shows much weaker skewness (the vorticity skewness
is 0.68 and 0.67 for hourly and daily-averaged velocities). The shape of the
vorticity-strain jPDF  appears to be a combination of two half-ellipses centered
about $\zeta=0$, one with a 45$^\circ$ slope (characteristic of submesocale fronts
that persist in Summer) and one with a very steep slope. The vorticity-divergence
jPDF is significantly different than the April one. In particular, the jPDF
for daily-averaged currents is still egg-shaped but hourly and residual fields are
ellipses about $\zeta=0$, likely associated with inertia-gravity waves.
That the submesoscale dynamics in October is mainly unbalanced is clearly depicted
in the shape of jPDF of vorticity-Laplacian of SSH for hourly and
residual fields
(figure \ref{fig3}p and \ref{fig3}r), which is an ellipse aligned in the vertical axis.
Of course daily-averaging the model suppresses the ageostrophic flows and
the daily-averaged flow is essentially geostrophic.

 Time series of vorticity and divergence PDFs (supplemental material) suggest
 a strong oscillation between these two regimes. In late Winter/early Spring
 the vorticity is strongly positively skewed, whereas the divergence is moderately
 negatively skewed as predicted by frontogenesis \citep{capet_etal2008a,mcwilliams2016}
. In late Summer/early Fall, the divergence is more stronger, but PDFs are much less skewed,
consistent with linear inertia-gravity waves.

@@

Figure \ref{fig4}a depicts the horizontal wavenumber spectra of surface KE in
April and October. At scales larger than 20km, the spectra based on hourly
velocity snapshots are nearly indistinguishable within 95$\%$ confidence level
(compare solid lines in figure \ref{fig3}a). Consistent with the results of
\citet{rocha_etal2016} who analyzed a 5-month output of the LLC4320 simulation in Drake Passage,
there is significant high-frequency variabilty at submesoscales. Daily-averaging
the velocity field suppresses spatial variability at scales smaller than about 250
km both in April and October (compare solid lines against dashed lines in figure
\ref{fig3}a). But this suppression is more dramatic in October, when the horizontally
divergent flows peak. At scales
smaller than 100 km, 39$\%$ of the surface KE in April is accounted for by super-inertial
flows as opposed to 79$\%$ in October. The seasonality of subinertial submesoscale flows
is more dramatic than the seasonal cycle of the total flow. This is consistent
with the results of \citet{sasaki_etal2014}, which are
based on daily-averaged velocity fields of a different model without tidal forcing
(P. Klein, personal communication).

These high-frequency submesoscale flows significantly project on the sea-surface.
In October, when the inertia-gravity waves are strongest the surface and submesoscale
turbulence is weakest, there is a dramatic difference between the spectra
based on hourly and daily-averaged SSH: at scales smaller than 100 km, the spectra
of hourly SSH roughly follows a $-2$ power-law, whereas the spectra of daily-averaged
SSH roughly follows a $-5$ power-law. At these scales, 33$\%$ of the SSH variance in April is accounted for by suprainertial
flows as opposed to 83$\%$ in October! Curiously, the out-of-phase seasonal cycle of submesoscale
turbulence and near-surface inertia-gravity waves conspire to yield very weak seasonality
in the spectra of KE and SSH variance based on hourly field.

@@

Our work adds to recent studies that presented modeling \cite{sasaki_etal2014}
and observational \citep{callies_etal2015,buckingham_etal2016} evidence of vigorous seasonality in
submesoscale turbulence. We conjecture that the dominance at inertia-gravity waves in
Summer \citep{callies_etal2015} is a consequence both of suppresion of submesoscale turbulence and
enhancement of inertia-gravity waves due to re-stratification of
the upper ocean. Of course, our study focuses on a single patch of ocean in the
vicinity of the Kuroshio Extension, which may be typical of mesoscale-rich
subtropical regions, but not
representative of other regions such as low eddy kinetic energy eastern boundary currents
and the middle of the subtropical gyre: care must be taken to avoid overgeneralizing
these results.  We plan to report on the geographic variability of submesoscale
 seasonality in a future study.

>>>>

\begin{abstract}
    Two new high-resolution numerical simulations with embedded tides suggest a
    strong modulation in the governing near-surface dynamics at submesoscales
    (roughly 10-100 km) in the Kuroshio Extension. Consistent with recent studies, deep late-Winter mixed layers
    are prone to baroclinic instabilities, and submesoscale turbulence
    prevails in late Winter/early Spring. Inertia-gravity waves are enhanced near
    the surface when the upper ocean is strongly stratified. Thus the summertime
    re-stratification weakens submesoscale turbulence and enhances intertia-gravity
     waves near the surface. In the Kuroshio Extension,
    inertia-gravity waves strongly dominate the surface kinetic energy and
    sea-surface height variance in late Summer/early Fall at submesoscales.
\end{abstract}

@@

\section{The LLC numerical simulations}
We use the output of two latitude-longitude polar cap (LLC)
realistic numerical simulations. The outputs
analyzed here, LLC2160 (1/24$^\circ$) and LLC4320 (1/48$^\circ$)  have
90-vertical-levels and are
forward Massachusetts Institute of Technology general circulation model (MITgcm)
numerical solutions on a LLC grid   \citep{forget_etal2015} that were
spun up from a ECCO2 adjoint-method state estimate. Both simulations were forced by
tides and 6-hourly surface fluxes. The LLC2160
output spans two years from March 2011 to April 2013; the LLC4320 was spun up from
the LLC2160 simulation, spanning one year from September 2011 to September 2012.
The LLC4320 simulation is an extension of the 5-month long output used by
\citet{rocha_etal2016}.

A key aspect of the LLC simulations is that they were forced by
the 16 most-significant tidal components.
Because barotropic tides interact with topography and generate internal
tides that project onto mesoscales to submesoscales
\citep[e.g., ][]{rocha_etal2016}, tidal forcing fundamentally distinguishes our analysis
from recent modeling studies that investigated upper-ocean dynamics
and its seasonality \citep{sasaki_etal2014,qiu_etal2014}. Details of the LLC simulations
are provided in the supplemental material.

To study seasonal variations in the upper-ocean dynamics, we focus on the northwest
Pacific, in the vicinity of the Kuroshio
Extension, where previous studies have suggested strong submesoscale seasonality
\citep{sasaki_etal2014,qiu_etal2014}.
We analyze a sub-domain of the LLC4320 and LLC2160 simulations of about 2000 km$^2$
spanning 155-175$^\circ$E; 25-40$^\circ$N (figure \ref{fig1}a). The stratification
in this mesoscale-rich subtropical region undergoes a vigorous seasonal cycle: wintertime-enhanced
small-scale turbulence de-stratifies the upper ocean, yielding mixed layers
as deep as 300m. In late Spring/early Summer enhanced solar radiation and mixed-layer
instabilities re-stratify the upper ocean, yielding mixed layers as shallow as 40m.
Fundamentally, the stratification and its seasonality are well-captured by both LLC simulations:
a comparison with Argo climatology shows that both simulations skillfully represent the Kuroshio
Extension stratification and its seasonal variability  (supplemental material).

@@

\section{Statistics of the surface lateral velocity tensor}
To study the seasonality in the surface velocity, we calculate the lateral velocity tensor
\begin{equation}
\nabla_h \bf{u}_h = \left[\begin{matrix} u_x & u_y\\ v_x&v_y \end{matrix}\right] \,.
\end{equation}
The components of the lateral velocity tensor are calculated using a centered
second-order finite difference scheme. We then diagnose
the vertical vorticity $\zeta \equiv v_x - u_y$, lateral rate of strain
$\alpha \equiv [(u_x-v_y)^2 + (u_y + v_x)^2]^{1/2}$, and horizontal divergence $\delta \equiv u_x + v_y$.
These diagnostics highlight the submesoscale structures in the flow
\citep[e.g.,][]{capet_etal2008a,shcherbina_etal2013}.

@@

The bulk of vertical vorticity and lateral strain rate are associated with subinertial flows
($T_f = 2\pi/f_0\approx 23.5$h, where $f_0$ is the inertial frequency at
the mean latitude): daily-averaging the velocity fields suppresses superinertial
motions and reduces the RMS
vorticity by less than 40$\%$ and the RMS strain by less than 10$\%$; the seasonal
cycle remains strong (see
dashed lines in figures \ref{fig2}a-b). Most of this seasonal cycle is associated
with submesoscale flows: smoothing the velocity fields
with a Hanning filter with cut-off scale of 100 km dramatically reduces the RMS
vorticity and strain. The reduction in variance is about 80$\%$ in winter,
yielding RMS vorticity and strain roughly consistent with the diagnostics from
AVISO gridded geostrophic velocities (compare red lines to black lines in figures \ref{fig2}a-b).
The overall picture that emerges is consistent with recent studies: shallow baroclinic
instabilities energize the submesoscales in late Winter, drawing from the available
potential energy stored in deep mixed layers \citep{sasaki_etal2014,callies_etal2015,callies_etal2016}.

The seasonal cycle of the horizontal divergence, however, showcases the complexity
of the upper-ocean annual variability. If submesoscale eddies and fronts dominated
the near-surface variability all year, then the seasonal cycle of horizontal divergence,
vertical vorticity, and lateral strain rate would be in phase \citep[e.g.,][]{sasaki_etal2014}.
While there is a clear wintertime peak in divergence of daily-averaged velocity
(see dashed lines in figure \ref{fig2}c; RMS divergence $\sim0.22 f$ in the 1/48$^\circ$
simulation), the hourly fields show
a stronger enhancement of
lateral divergence in late Summer/early Fall (RMS divergence $\sim0.22 f$ the 1/48$^\circ$
simulation). Because the 1/48$^\circ$ simulation better resolves
smaller-scale submesoscale flows, a secondary RMS divergence peak in winter  is
nearly as strong as in Summer. Submesoscale fronts and eddies evolve
relatively fast, and there is no clear
temporal and spatial scale separation between those motions and inertia-gravity waves
\citep{mcwilliams2016}:
daily-averaging the velocity fields efficiently suppresses the summertime horizontally
divergent flows, but also reduces the wintertime lateral divergence by about 50$\%$.
The vorticity and strain rate (Figure \ref{fig2}a-b) show that most of the lateral divergence is associated
with submesoscale flows: smoothing the velocity fields with a 100km-cut-off
suppresses more than 80$\%$ of the RMS divergence.

\section{Seasonality in governing submesoscale dynamics}
The overall results of figure \ref{fig2}c suggest that submesoscale surface
 variability stems from different dynamics in Summer than Winter.
 To better understand these differences, we calculate
joint probability distributions (jPDF) of
vorticity-strain and vorticity-Laplacian of sea-surface height (\ref{fig3}).

The April vorticity-strain jPDF has a shape characteristic of submesoscale turbulence
 (see figures \ref{fig3}a-c). The alignment of vorticity and strain $\alpha \sim \pm\zeta$
 with strong positive skewness are fingerprints of submesoscale fronts
 \citep{shcherbina_etal2013,mcwilliams2016}. The shape of the vorticity-strain
 jPDF is similar for hourly, daily-averaged, and residual velocities, although
 the vorticity skewness reduces 1.4 to 1.13 from hourly to daily-averaged.
 Thus wintertime submesoscale surface velocity is strongly dominated by submesoscale turbulence --- this is true even for
 superinertial submesoscale currents. The hourly velocity fields are significantly
ageostrophic as depicted by the jPDF of vorticity-Laplacian of SSH
(figure \ref{fig3}g). While there is certainly much to be
learned using quasi-geostrophic models to study submesoscale turbulence, claims that
the submesoscale turbulence is dominantly in geostrophic balance
are overstated \citep{sasaki_etal2014,callies_etal2016}; only the daily-averaged fields
are largely
in  geostrophic balance (the jPDF of daily-averaged vorticity vs. Laplacian of SSH
is an ellipse with large eccentricity with main axis tilted by 45$^\circ$; figure \ref{fig3}h).

The October vorticity-strain jPDF shows much weaker skewness (the vorticity skewness
is 0.68 and 0.67 for hourly and daily-averaged velocities). The shape of the
vorticity-strain jPDF  appears to be a combination of two half-ellipses centered
about $\zeta=0$, one with a 45$^\circ$ slope (characteristic of submesocale fronts
that persist in Summer) and one with a very steep slope.
That the submesoscale dynamics in October is mainly unbalanced is clearly depicted
in the shape of jPDF of vorticity-Laplacian of SSH for hourly and
residual fields
(figure \ref{fig3}j and \ref{fig3}k), which is an ellipse aligned in the vertical axis.
Of course daily-averaging the model suppresses the ageostrophic flows, and
the daily-averaged flow is essentially geostrophic as depicted by the 45$^\circ$-tilted
ellipse in the vorticity-Laplacian of SSH jPDF.

 Time series of vorticity and divergence PDFs (supplemental material) suggest
 a strong oscillation between these two regimes. In late Winter/early Spring
 the vorticity is strongly positively skewed, whereas the divergence is moderately
 negatively skewed as predicted by frontogenesis \citep{capet_etal2008a,mcwilliams2016}
. In late Summer/early Fall, the divergence is stronger, but PDFs are much less skewed,
consistent with linear inertia-gravity waves.

@@

Figure \ref{fig4}a depicts the horizontal wavenumber spectra of surface KE in
April and October. At scales larger than 20km, the spectra based on hourly
velocity snapshots are nearly indistinguishable within 95$\%$ confidence level
(compare solid lines in figure \ref{fig4}a). Consistent with the results of
\citet{rocha_etal2016} who analyzed a 5-month output of the LLC4320 simulation in Drake Passage,
there is significant high-frequency variabilty at submesoscales. Daily-averaging
the velocity field suppresses spatial variability at scales smaller than about 250
km both in April and October (compare solid lines against dashed lines in figure
\ref{fig4}a). But this suppression is dramatic in October, when the horizontally
divergent flows peak. At scales
smaller than 100 km, 39$\%$ of the surface KE in April is accounted for by super-inertial
flows as opposed to 79$\%$ in October. The seasonality of subinertial submesoscale flows
is more dramatic than the seasonal cycle of the total flow. This is consistent
with the results of \citet{sasaki_etal2014}, which are
based on daily-averaged velocity fields of a different model without tidal forcing
(P. Klein, personal communication).

These high-frequency submesoscale flows significantly project on the sea-surface.
In October, when the inertia-gravity waves are strongest the surface and submesoscale
turbulence is weakest. There is a dramatic difference between the spectra
based on hourly and daily-averaged SSH: at scales smaller than 100 km, the spectra
of hourly SSH roughly follows a $-2$ power-law, whereas the spectra of daily-averaged
SSH roughly follows a $-5$ power-law. At these scales, 33$\%$ of the SSH variance in April is accounted for by suprainertial
flows as opposed to 83$\%$ in October! Curiously, the out-of-phase seasonal cycle of submesoscale
turbulence and near-surface inertia-gravity waves conspire to yield very weak seasonality
in the spectra of KE and SSH variance based on hourly fields.

@@

Our work adds to recent studies that presented modeling \cite{sasaki_etal2014}
and observational \citep{callies_etal2015,buckingham_etal2016} evidence of vigorous seasonality in
submesoscale turbulence. We conjecture that the summertime dominance of inertia-gravity waves
 \citep{callies_etal2015} is a consequence both of suppression of submesoscale turbulence and
enhancement of inertia-gravity waves due to re-stratification of
the upper ocean. Of course, our study focuses on a single patch of ocean in the
vicinity of the Kuroshio Extension, which may be typical of mesoscale-rich
subtropical regions, but not
representative of other regions such as low eddy kinetic energy eastern boundary currents
and the middle of the subtropical gyre; care must be taken to avoid overgeneralizing
these results.  We plan to report on the geographic variability of submesoscale
 seasonality in a future study.
